---
title: "Working with Data"
author: "Alessandro Barberis"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: false
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Objective

In this lab we will work with a real clinical dataset from a prostate cancer cohort.
By the end of the session you will be able to:

- import a tabular dataset into R
- explore its structure and contents
- create basic visualisations with [`ggplot2`](https://ggplot2.tidyverse.org)
- save your plots to disk

We introduce [`ggplot2`](https://ggplot2.tidyverse.org) here as a practical tool. In later labs we will use it
extensively, so it is worth getting comfortable with its core syntax now.

---

# Setup

## Project structure

This lab assumes you are working inside the course [RStudio](https://posit.co/download/rstudio-desktop/) project. The project uses
the [`here`](https://here.r-lib.org/) package to construct file paths relative to the project root, so the code
works regardless of where you have saved the project on your computer.

If you do not have `here` or [`ggplot2`](https://ggplot2.tidyverse.org) installed, run the chunk below **once**:

```{r install, eval=FALSE}
install.packages(c("here", "ggplot2"))
```

```{r libraries}
library(here)
library(ggplot2)
```

## The `here` package

`here::here()` always returns the path to the project root, no matter which
subfolder your `.qmd` file is in. For example:

```{r here_example}
# Shows the project root on your machine
here::here()
```

All file paths in this document are built with `here()`, so you never need to
adjust them manually.

---

# Dataset: TCGA Prostate Cancer

## About the data

We use clinical data from the **Prostate Adenocarcinoma** cohort of [The Cancer Genome
Atlas (TCGA)](https://www.cancer.gov/ccg/research/genome-sequencing/tcga), as curated in the
[TCGA PanCancer Atlas 2018](https://www.cbioportal.org/study/summary?id=prad_tcga_pan_can_atlas_2018).
This cohort includes 499 patients with matched genomic and clinical data.

[**cBioPortal**](https://www.cbioportal.org) is an open-access platform that provides interactive exploration of
large-scale cancer genomics datasets. It is widely used in translational oncology
research and is a key resource you will encounter throughout this master.

## Download the data

1. Go to the
   [TCGA Prostate Cancer study page](https://www.cbioportal.org/study/summary?id=prad_tcga_pan_can_atlas_2018)
   on cBioPortal.
2. Click **Download**.
3. Unzip the downloaded archive and move the folder
   `prad_tcga_pan_can_atlas_2018/` into the `data/` folder of the course project.

The file we need is `data_clinical_patient.txt` inside that folder.

::: {.callout-note}
The dataset is already available in the `data/` folder if you downloaded the full
course materials from GitHub.
:::

---

# Import data

The clinical file uses tab-separated values with a 4-line header that describes the
columns. We skip those lines and read the actual data starting at line 5.

```{r import}
clinical_data <- read.csv(
  file             = here("data", "prad_tcga_pan_can_atlas_2018", "data_clinical_patient.txt"),
  header           = TRUE,
  skip             = 4,
  sep              = "\t",
  stringsAsFactors = FALSE
)
```

---

# Explore data

Before plotting or modelling, always explore your data. A few quick checks can save
hours of debugging later.

```{r explore_head}
# Preview the first 6 rows
head(clinical_data)
```

```{r explore_str}
# Column names, types, and a sample of values
str(clinical_data)
```

```{r explore_summary}
# Statistical summary of each column
summary(clinical_data)
```

::: {.callout-tip}
## What to look for
When exploring a new dataset, pay attention to:

- **Missing values** (`NA`): `summary()` reports their count per column.
- **Unexpected types**: a numeric column stored as character is a common import issue.
- **Implausible values**: e.g. a patient age of 0 or 999 usually signals a coding
  convention for missing data.
:::

---

# Visualise data with `ggplot2`

[`ggplot2`](https://ggplot2.tidyverse.org) implements the *Grammar of Graphics*: every plot is built by combining a
**dataset**, an **aesthetic mapping** (which variables map to x, y, colour, etc.),
and one or more **geometric layers** (points, lines, bars, etc.).

The basic template is always:

```r
ggplot(data, aes(x = ..., y = ...)) +
  geom_*() +
  labs(...)
```

## Scatter plot: age vs. follow-up duration

```{r scatter}
ggplot(clinical_data, aes(x = AGE, y = DAYS_LAST_FOLLOWUP)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Age at diagnosis vs. days to last follow-up",
    x     = "Age at diagnosis (years)",
    y     = "Days to last follow-up"
  )
```

This plot shows the relationship between age at diagnosis and the length of
follow-up. Each point is one patient. The `alpha` argument controls point
transparency, which helps reveal overplotted regions.

---

# Save a plot

Use `ggsave()` to export the last plot to a file. Specifying `width`, `height`,
and `dpi` ensures consistent, publication-quality output.

```{r save_plot, eval=FALSE}
ggsave(
  filename = here("output", "scatter_age_followup.png"),
  plot     = last_plot(),
  width    = 8,
  height   = 6,
  dpi      = 300
)
```

::: {.callout-note}
The `output/` folder must exist before running this chunk. Create it in the Files
pane of RStudio, or run `dir.create(here("output"))` once.
:::

---

# Exercises

Work through the exercises below. Try to write the code yourself before looking at
the solution.

## Exercise 1: Distribution of age at diagnosis

Create a histogram showing the distribution of patient age at diagnosis.

**Hint:** look at the [`ggplot2`](https://ggplot2.tidyverse.org) documentation for `geom_histogram()`.
Try adjusting the `binwidth` argument to see how it affects the plot.

```{r ex1, eval=FALSE}
# Your code here

```

::: {.callout-tip collapse="true"}
## Solution

```{r ex1-solution}
ggplot(clinical_data, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(
    title = "Distribution of age at diagnosis",
    x     = "Age at diagnosis (years)",
    y     = "Number of patients"
  )
```
:::

## Exercise 2: Overall survival by pathological stage

Create a violin plot showing the distribution of overall survival time (in months)
for each pathological T-stage.

**Hints:**

- Look at `geom_violin()` in the `ggplot2` documentation.
- Add a `geom_jitter()` layer on top to show individual data points.
- The relevant columns are `PATH_T_STAGE` and `OS_MONTHS`.

```{r ex2, eval=FALSE}
# Your code here

```

::: {.callout-tip collapse="true"}
## Solution

```{r ex2-solution}
ggplot(clinical_data, aes(x = PATH_T_STAGE, y = OS_MONTHS)) +
  geom_violin(fill = "lightblue", alpha = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 0.8) +
  labs(
    title = "Overall survival by pathological T-stage",
    x     = "Pathological T-stage",
    y     = "Overall survival (months)"
  )
```

Notice how the violin plot reveals the shape of the distribution within each stage,
while the jittered points show the actual density of observations.
:::

## Exercise 3 (optional): Kaplan–Meier survival curves

::: {.callout-note collapse="true"}
## Going deeper: survival analysis

If you are familiar with survival analysis, try plotting Kaplan–Meier curves
stratified by pathological stage using the `survminer` and `survival` packages.

```{r ex3-solution, eval=FALSE}
# install.packages(c("survival", "survminer"))
library(survival)
library(survminer)

# Create a survival object
# OS_STATUS coding: check the data dictionary — typically 0 = alive, 1 = dead
surv_obj <- Surv(time = clinical_data$OS_MONTHS,
                 event = clinical_data$OS_STATUS == "DECEASED")

# Fit KM curves by pathological stage
km_fit <- survfit(surv_obj ~ PATH_T_STAGE, data = clinical_data)

# Plot
ggsurvplot(
  km_fit,
  data       = clinical_data,
  risk.table = TRUE,
  pval       = TRUE,
  title      = "Kaplan–Meier survival by pathological T-stage"
)
```

We will cover survival analysis in more detail in later sessions.
:::

---

# Session Info

```{r session_info}
sessionInfo()
```
