---
title: "Introduction to R"
author: "Alessandro Barberis"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: false
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Markdown and Quarto

This document is written in **Quarto**, the next-generation version of R Markdown.
Quarto documents mix prose, code, and output in a single file, and can be rendered to
HTML, PDF, or Word. You are reading the HTML output.

Each grey box below is an **R code chunk**. When you render the document, R executes
each chunk in order and embeds the output directly beneath it. You can also run chunks
interactively in RStudio by clicking the green arrow on the top-right of each chunk.

```{r example-chunk}
# This is a code chunk. Run it to see the output below.
summary(cars)
```

::: {.callout-tip}
## How to use this document
The sections marked **Core** cover the material we will go through together in the
lab session. Sections inside collapsible boxes marked **Going deeper** are additional
material for you to explore at your own pace after the session.
:::

---

# Data types

Every value in R has a type. Understanding types is essential because many common
errors in R arise from a mismatch between the type you expect and the type you actually
have (for instance, a number stored as text that refuses to be plotted on a numeric axis).

The most common types you will encounter in biomedical data analysis are **logical**,
**integer**, **double**, and **character**. We cover these first, then provide an
overview of the less common types for reference.

You can always inspect any object with three complementary functions:

| Function | What it tells you |
|----------|-------------------|
| `typeof()` | Internal storage type |
| `mode()` | Storage mode (similar to `typeof`, slightly higher level) |
| `class()` | How R treats the object in function dispatch |

## Logical

Logical values are `TRUE` or `FALSE`. They arise naturally from comparisons and
conditions, and are the backbone of filtering and subsetting data.

```{r logical_1}
# Create a logical vector
l <- c(TRUE, FALSE, TRUE)
l
```

```{r logical_2}
typeof(l)
mode(l)
class(l)
```

## Integer

Integers are whole numbers, stored more efficiently than decimals. In R they are
written with a trailing `L`. You will encounter them most often as counts or indices.

```{r integer_1}
# Create an integer vector
i <- c(1L, 2L, 3L)
i
```

```{r integer_2}
typeof(i)
mode(i)
class(i)
```

## Double

Doubles (also called *numeric* in R) are real numbers with decimal precision. Most
continuous measurements — age, weight, expression values, survival time — are doubles.

```{r double_1}
# Create a double vector
d <- c(1.0, 2.5, 3.7)
d
```

```{r double_2}
typeof(d)
mode(d)
class(d)
```

## Character

Character values (strings) hold text. They are used for labels, patient IDs, gene
names, and any categorical variable before it is converted to a factor.

```{r character_1}
# Create a character vector
c.v <- c("a", "b", "c")
c.v
```

```{r character_2}
typeof(c.v)
mode(c.v)
class(c.v)
```

## Less common types {.unnumbered}

::: {.callout-note collapse="true"}
## Going deeper: complex and raw types

These two types rarely appear in biomedical data analysis, but are included here for
completeness.

**Complex** numbers have a real and an imaginary part (written `a + bi`). They arise
in signal processing and certain mathematical transformations (e.g. Fourier analysis).

```{r complex_1}
cpx <- c(1+2i, 3+4i, 5+6i)
cpx
typeof(cpx)
```

**Raw** values store binary data as bytes. You might encounter them when reading binary
file formats or network data.

```{r raw_1}
r <- as.raw(c(1, 2, 3))
r
typeof(r)
```
:::

---

# Data structures

A data structure is a way of organising multiple values. The four structures you will
use most in data analysis are **vectors**, **matrices**, **data frames**, and **lists**.
The table below gives a quick orientation:

| Structure | Dimensions | Homogeneous? | Typical use |
|-----------|-----------|--------------|-------------|
| Vector | 1D | Yes | Single variable |
| Matrix | 2D | Yes | Numeric data (expression matrix) |
| List | 1D | No | Mixed-type collections |
| Data frame | 2D | No (per column) | Clinical / omics datasets |

## Vectors

A vector is a one-dimensional, ordered collection of values **all of the same type**.
It is the fundamental building block in R — a single number is actually a vector of
length 1.

### Create

```{r vector_1}
# Combine values with c()
v <- c(1, 2, 3, 4, 5)
v
```

```{r vector_2}
# Regular sequence with seq()
v <- seq(from = 1, to = 10, by = 2)
v
```

```{r vector_3}
# Repeated pattern with rep()
v <- rep(x = 1:3, times = 2)
v
```

```{r vector_4}
# Random sample (set.seed() makes the result reproducible)
set.seed(42)
v <- sample(x = 1:10, size = 5, replace = FALSE)
v
```

::: {.callout-tip}
## Reproducibility tip
Use `set.seed()` before any function that involves randomness. This ensures that
anyone running your code will get the same result — essential for reproducible research.
:::

### Inspect

```{r vector_5}
typeof(v)
class(v)
length(v)
```

Elements are accessed with square bracket indexing `[]`. R uses **1-based indexing**
(the first element is `[1]`, not `[0]` as in Python).

```{r vector_subsetting}
v[1]        # first element
v[1:3]      # elements 1 to 3
```

Vectors can also have names, allowing access by label:

```{r vector_subsetting3}
names(v) <- c("a", "b", "c", "d", "e")
v["a"]
```

### Modify

```{r vector_modify1}
v <- c(v, 6)         # append an element
v
```

```{r vector_modify2}
v <- v[-1]           # remove first element
v
```

```{r vector_modify3}
v[1] <- 10           # overwrite first element
v
```

## Matrices

A matrix is a two-dimensional, homogeneous structure — essentially a vector with
rows and columns. Matrices are common in bioinformatics: gene expression data is
often stored as a matrix with genes on rows and samples on columns.

### Create

```{r matrix_1}
m <- matrix(
  data  = 1:9,
  nrow  = 3,
  ncol  = 3,
  byrow = FALSE    # fill column by column (default)
)
m
```

A matrix can also be created by assigning dimensions to a vector with `dim()`:

```{r matrix_from_vector2}
v2 <- 1:9
dim(v2) <- c(3, 3)
identical(m, v2)   # TRUE: the two objects are the same
```

### Inspect

Elements are indexed by `[row, col]`. Leaving one dimension blank selects everything
in that dimension.

```{r matrix_subsetting}
m[2, 1]          # element at row 2, column 1
m[1:2, ]         # first two rows, all columns
```

Named indexing also works for matrices:

```{r matrix_subsetting3}
rownames(m) <- c("a", "b", "c")
colnames(m) <- c("d", "e", "f")
m["a", "d"]
```

```{r matrix_dim}
dim(m)     # returns c(nrows, ncols)
```

### Modify

```{r matrix_modify}
m <- rbind(m, c(10, 11, 12))   # add a row
m <- m[-4, ]                   # remove 4th row
m <- cbind(m, c(13, 14, 15))   # add a column
m <- m[, -4]                   # remove 4th column
m[1, 1] <- 10                  # modify one element
m
```

## Lists

A list is a one-dimensional, **heterogeneous** structure: each element can be of any
type, including another list. Lists are how R represents complex objects such as the
output of a statistical model.

### Create

```{r list_1}
l <- list(
  name   = "John",
  age    = 30,
  height = 1.75
)
l
```

### Inspect

Lists support three types of subsetting, with subtly different results:

```{r list_subsetting}
l[1]       # returns a list of length 1
l[[1]]     # returns the element itself
l$name     # same as l[["name"]], using the element name
```

```{r list_length}
length(l)
```

### Modify

```{r list_modify}
l <- c(l, list(weight = 70))   # add an element
l$height <- NULL               # remove an element
l$name <- "Jane"               # modify an element
l
```

::: {.callout-note collapse="true"}
## Going deeper: the `[[` vs `[` distinction

`[` always returns a list (the container), while `[[` returns the element inside.
This matters when passing list elements to functions:

```{r list_double_bracket}
# These two calls behave differently
class(l[1])    # "list"
class(l[[1]])  # "character"
```

Most functions expect the element itself, not a list containing it — so `[[` and `$`
are more commonly useful when working with list contents.
:::

## Data frames

A data frame is the central data structure for tabular data in R. It is a
**two-dimensional, heterogeneous** structure: a collection of vectors of equal length,
each potentially of a different type. Each row represents an observation (e.g. a
patient) and each column a variable (e.g. age, sex, diagnosis).

### Create

```{r dataframe_1}
df <- data.frame(
  patientid = 1:3,
  sex       = c("M", "F", "M"),
  age       = c(34, 41, 57)
)
df
```

### Inspect

Three functions cover most inspection needs:

```{r dataframe_inspect}
head(df)       # first 6 rows
str(df)        # structure: column types and a preview of values
summary(df)    # statistical summary per column
```

### Modify

```{r dataframe_modify}
# Add a column
df$height <- c(1.75, 1.68, 1.80)
df

# Remove a column
df$height <- NULL
df

# Add a row — use data.frame() to preserve types
df <- rbind(df, data.frame(patientid = 4, sex = "F", age = 51))
df
```

::: {.callout-warning}
## Type coercion trap
Avoid using `c()` to create new rows for a data frame. Because `c()` coerces
everything to the most general type in the vector, numeric columns will be silently
converted to character:

```{r coercion_trap}
# This coerces patientid and age to character — probably not what you want
df_bad <- rbind(df, c(5, "M", 28))
str(df_bad)
```
:::

## Factors

Factors represent **categorical variables** with a fixed set of possible values
(called *levels*). They are important for statistical modelling — if you include
sex or tumour stage in a regression, R needs to know they are categories, not
arbitrary numbers or strings.

### Create

```{r factor_1}
f <- factor(
  c("M", "F", "M"),
  levels = c("M", "F")
)
f
```

### Inspect

```{r factor_inspect}
levels(f)       # the possible categories
table(f)        # frequency of each level
summary(f)      # similar to table() for factors
str(f)          # internal representation (integers + levels)
```

::: {.callout-warning}
## `as.numeric()` on factors
`as.numeric()` returns the underlying integer codes (1, 2, 3...), **not** the level
labels. This is a common source of errors. To get numeric values from a factor whose
labels are numbers, use `as.numeric(as.character(f))`.

```{r factor_numeric}
f2 <- factor(c("10", "20", "30"))
as.numeric(f2)              # returns 1 2 3 — WRONG
as.numeric(as.character(f2)) # returns 10 20 30 — correct
```
:::

### Modify

```{r factor_modify}
levels(f) <- c(levels(f), "U")   # add a level
f <- droplevels(f)               # remove unused levels
f
```

---

# Functions

A function is a reusable block of code that takes inputs (arguments) and returns an
output. R comes with thousands of built-in functions, and writing your own is essential
for keeping your analysis organised and avoiding repetition.

## Create

```{r fun_1}
myfun <- function(x, y) {
  result <- x + y
  return(result)
}
```

## Inspect

```{r fun_inspect}
formals(myfun)      # the argument list
body(myfun)         # the code inside
```

## Call

Arguments can be passed by **position** or by **name**. Using names makes code more
readable and prevents errors when a function has many arguments.

```{r fun_call}
myfun(4, 6)             # by position
myfun(x = 4, y = 6)    # by name (preferred)
```

---

# Input / Output

In practice you will almost always read data from a file and save results to disk.

## Save and load R objects

`save()` and `load()` preserve R objects in their native binary format (`.RData`).
This is useful for caching intermediate results in a long analysis.

```{r save_load, eval=FALSE}
save(df, file = "data.RData")
load(file = "data.RData")
```

## Read and write CSV

CSV is the most portable format for sharing data between R and other tools.

```{r read_write_csv, eval=FALSE}
# Write
write.csv(df, file = "data.csv", row.names = FALSE)

# Read
df <- read.csv(file = "data.csv")
```

::: {.callout-note collapse="true"}
## Going deeper: the `readr` and `data.table` packages

Base R's `read.csv()` is fine for small files. For larger datasets (hundreds of
thousands of rows), consider:

- `readr::read_csv()` from the tidyverse: faster than base R, better type guessing,
  and returns a *tibble* instead of a plain data frame.
- `data.table::fread()`: extremely fast for very large files.

Both packages need to be installed once with `install.packages()` before use.
:::

---

# Session Info

The section below records the exact versions of R and all loaded packages used to
generate this document. Including it at the end of every analysis script is good
practice for reproducibility.

```{r session_info}
sessionInfo()
```
